Methods paper for erosion

### 2017-11-08 docs/methods/erosion_methods.tex
minior latex changes to paper

\author[Nick]{M.~Nicholas J.~Moore}
\author[Bryan]{Bryan D.~Quaife}
\address[Nick]{Department of Mathematics and Geophysical Fluid Dynamics Institute, Florida State University, Tallahassee, FL, 32306.}
\address[Bryan]{Department of Scientific Computing and Geophysical Fluid Dynamics Institute, Florida State University, Tallahassee, FL, 32306.}

@@

\todo[inline]{Bryan, I think you should be first author on this paper.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
\label{sec:spatial_discretization}
Each body is represented with $N_\iin$ points as the Fourier series
\begin{align*}
  \xx^k(\theta) = \sum_{j = -N_\iin /2}^{N_\iin /2-1} \hat{x}_j^k e^{ij\theta},
    \quad k=1,\ldots,M.
\end{align*}
The exterior geometry is represented with a Fourier series using
$N_\out$ points.  Derivatives of the shape are computed in Fourier space
\begin{align*}
  \frac{d\xx^k(\theta)}{d\theta} = 
    \sum_{j = -N_\out /2}^{N_\out /2-1} ij\hat{x}_j^k e^{ij\theta},
    \quad k=1,\ldots,M.
\end{align*}

@@

In order to regularize boundary evolution, we modify the normal velocity
in Eq.~(\ref{eqn:bodyEvolution}) as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \frac{\eps}{\log \left(2\pi/L \right)} \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body. The second term on the right acts to smooth bodies that have high-frequency oscillations. Since we have subtracted the mean curvature, $2\pi/L$, this term is area-preserving, so that the only material loss is due to the shear stress. We chose the prefactor $\log^{-1} \left(2\pi/L \right)$, so that, for 2D Stokes flow, this term scales the same as shear stress as $L \to 0$. Thus, the relative strength of the smoothing is always the same.

We now introduce the tangent angle, $\theta$, defined through the relation
\begin{equation}
\left( \pderiv{x}{s}, \pderiv{y}{s} \right) = \left(\cos \theta, \sin \theta \right)
\end{equation}
where $s$ is arc length. We also introduce the normalized arc length $\alpha = s / L$. In terms of these new variables, the curvature can be expressed as
\begin{equation}
\kappa = \frac{1}{L} \pderiv{\theta}{\alpha}
\end{equation}
Substitution into Eq.~(\ref{eqn:Vn}) gives
\begin{equation}
\label{eqn:Vn2}
\Vn = \atau + \frac{\eps}{L \log \left(2\pi/L \right)}  \left(\pderiv{\theta}{\alpha} - 2 \pi \right)
\end{equation}

The main idea is to introduce an artificial tangential velocity that keeps collocation points equally spaced with respect to arc length; i.e.~the grid of $\alpha$ should always remain equally spaced. The tangential velocity that does the trick is determined by \cite{hou-low-she1994}
\begin{equation}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}
\end{equation}
The interface evolution is then given by
\begin{equation}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{equation}
Upon transforming to the variables $\theta(\alpha,t)$ and $L(t)$, we obtain the evolution equations
\begin{align}
\label{eqn:Levo0}
& \tderiv{L}{t} = - \mean{\thalpha \Vn} \\
\label{eqn:thetaevo0}
& \pderiv{\theta}{t} = \frac{1}{L} \left( \pderiv{\Vn}{\alpha} + \thalpha \Vs \right)
\end{align}
Inserting Eq.~(\ref{eqn:Vn2}) for the normal velocity gives
\begin{equation}
\pderiv{\theta}{t} = \frac{\eps}{L^{2} \log \left(2\pi/L \right)} \ppd{\theta}{\alpha} 
+ \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right)
\end{equation}

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& M = - \mean{\thalpha \Vn} \\
\label{eq:N}
& N = \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
& \elfun = \frac{1}{L^{2} \log \left(2\pi/L \right)}
\end{align} 
Then Eq.~(\ref{eqn:thetaevo0}) can be rewritten as
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{equation}
Note that the $\elfun$ depends only on $L(t)$, and so the highest order derivative in Eq.~(\ref{eqn:thetaevo1}) is {\em linear} in the variable $\alpha$. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for the stiff term. We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \sum_{k = -\infty}^{\infty} \thhat_k e^{2\pi i k \alpha} \\
& N(\alpha,t)  = \sum_{k = -\infty}^{\infty} \hat{N}_k e^{2\pi i k \alpha}
\end{align}
Equation (\ref{eqn:thetaevo1}) gives a set of ODEs for the coefficients
\begin{equation}
\tderiv{\thhat_k}{t} +  4 \eps \pi^2 k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( 4 \eps \pi^2 k^2 \int \elfun \, dt \right)
\end{equation}
Then the evolution equations for $L$ and $\theta$ can be recast as the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = M \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \theta_k \right) = \mu_k \hat{N}_k
\end{align}
We solve system (\ref{eqn:Levo2})--(\ref{eqn:thetaevo2}) for the variables $L$ and $\{ \mu_k \theta_k \}$ with a 2nd-order Runge-Kutta method. At each stage, we must recover $\theta(\alpha,t)$ from  $\{ \mu_k \theta_k \}$ in order to compute the shear stress as needed in Eqs.~(\ref{eq:M}) and (\ref{eq:N}) for $M$ and $N$. For this, we compute $\mu_k$ from Eq.~(\ref{eq:mu}) using the trapezoid rule, then recover $\theta(\alpha, t)$ via a Gaussian filter....

>>>>

\author[Bryan]{Bryan D.~Quaife}
\author[Nick]{M.~Nicholas J.~Moore}
\address[Nick]{Department of Mathematics and Geophysical Fluid Dynamics Institute, Florida State University, Tallahassee, FL, 32306.}
\address[Bryan]{Department of Scientific Computing and Geophysical Fluid Dynamics Institute, Florida State University, Tallahassee, FL, 32306.}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
\label{sec:spatial_discretization}
Each body is represented with $N_\iin$ points as the Fourier series
\begin{align*}
  \xx^k(\alpha) = \sum_{j = -N_\iin /2}^{N_\iin /2-1} 
    \hat{x}_j^k e^{ij\alpha}, \quad k=1,\ldots,M.
\end{align*}
The exterior geometry is represented with a Fourier series using
$N_\out$ points.  Derivatives of the shape are computed in Fourier space
\begin{align*}
  \frac{d\xx^k(\alpha)}{d\alpha} = 
    \sum_{j = -N_\out /2}^{N_\out /2-1} ij\hat{x}_j^k e^{ij\alpha},
    \quad k=1,\ldots,M.
\end{align*}

@@

In order to regularize boundary evolution, we modify the normal velocity
in Eq.~\eqref{eqn:bodyEvolution} as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \frac{\eps}{\log \left(2\pi/L \right)} \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body. The second term on the right acts to smooth bodies that have high-frequency oscillations. Since we have subtracted the mean curvature, $2\pi/L$, this term is area-preserving, so that the only material loss is due to the shear stress. We chose the prefactor $\log^{-1} \left(2\pi/L \right)$, so that, for 2D Stokes flow, this term scales the same as shear stress as $L \to 0$. Thus, the relative strength of the smoothing is always the same.

We now introduce the tangent angle, $\theta$, defined through the relation
\begin{equation}
\left( \pderiv{x}{s}, \pderiv{y}{s} \right) = \left(\cos \theta, \sin \theta \right)
\end{equation}
where $s$ is arc length. We also introduce the normalized arc length $\alpha = s / L$. In terms of these new variables, the curvature can be expressed as
\begin{equation}
\kappa = \frac{1}{L} \pderiv{\theta}{\alpha}
\end{equation}
Substitution into Eq.~\eqref{eqn:Vn} gives
\begin{equation}
\label{eqn:Vn2}
\Vn = \atau + \frac{\eps}{L \log \left(2\pi/L \right)}  \left(\pderiv{\theta}{\alpha} - 2 \pi \right)
\end{equation}

The main idea is to introduce an artificial tangential velocity that keeps collocation points equally spaced with respect to arc length; i.e.~the grid of $\alpha$ should always remain equally spaced. The tangential velocity that does the trick is determined by \cite{hou-low-she1994}
\begin{equation}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}
\end{equation}
The interface evolution is then given by
\begin{equation}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{equation}
Upon transforming to the variables $\theta(\alpha,t)$ and $L(t)$, we obtain the evolution equations
\begin{align}
\label{eqn:Levo0}
& \tderiv{L}{t} = - \mean{\thalpha \Vn} \\
\label{eqn:thetaevo0}
& \pderiv{\theta}{t} = \frac{1}{L} \left( \pderiv{\Vn}{\alpha} + \thalpha \Vs \right)
\end{align}
Inserting Eq.~\eqref{eqn:Vn2} for the normal velocity gives
\begin{equation}
\pderiv{\theta}{t} = \frac{\eps}{L^{2} \log \left(2\pi/L \right)} \ppd{\theta}{\alpha} 
+ \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right)
\end{equation}

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& M = - \mean{\thalpha \Vn} \\
\label{eq:N}
& N = \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
& \elfun = \frac{1}{L^{2} \log \left(2\pi/L \right)}
\end{align} 
Then Eq.~\eqref{eqn:thetaevo0} can be rewritten as
\begin{equation}
\label{eqn:thetaevo1}
\pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{equation}
Note that the $\elfun$ depends only on $L(t)$, and so the highest order
derivative in Eq.~\eqref{eqn:thetaevo1} is {\em linear} in the variable $\alpha$. This is the key to the {\thL} method, and allows us to use very stable, implicit methods for the stiff term. We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \sum_{k = -\infty}^{\infty} \thhat_k e^{2\pi i k \alpha} \\
& N(\alpha,t)  = \sum_{k = -\infty}^{\infty} \hat{N}_k e^{2\pi i k \alpha}
\end{align}
\todo[inline]{Earlier, I assumed the geometry is $2\pi$ periodic.  Is it
possible to use this convention cleanly for the presentation?  This
allows us to avoid the long superscripts $2\pi i k \alpha$}
Equation~\eqref{eqn:thetaevo1} gives a set of ODEs for the coefficients
\begin{equation}
\tderiv{\thhat_k}{t} +  4 \eps \pi^2 k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( 4 \eps \pi^2 k^2 \int \elfun \, dt \right)
\end{equation}
Then the evolution equations for $L$ and $\theta$ can be recast as the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = M \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \theta_k \right) = \mu_k \hat{N}_k
\end{align}
We solve system~\eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} for the
variables $L$ and $\{ \mu_k \theta_k \}$ with a 2nd-order Runge-Kutta
method. At each stage, we must recover $\theta(\alpha,t)$ from  $\{
  \mu_k \theta_k \}$ in order to compute the shear stress as needed in
  Eqs.~\eqref{eq:M} and~\eqref{eq:N} for $M$ and $N$. For this, we
  compute $\mu_k$ from Eq.~\eqref{eq:mu} using the trapezoid rule, then recover $\theta(\alpha, t)$ via a Gaussian filter....
